---
title: "econometrics_project"
output: html_document
---

SECTION 1
```{r}
library (carData)
library(dplyr)
library(MASS)
library(readxl)
```
STEP 1
```{r}
first_data <- Wong
help(Wong)
first_data$age <- as.integer(first_data$age)

#cutting the data where duration of being in coma(in days) is greater than zero and where IQ was measured in 100 days after way out of coma
wong1 <- subset(first_data, first_data$duration>0)
wong2 <- subset(wong1, wong1$days < 100)

plot(wong2$viq, wong2$piq, xlab = "Vocabulary IQ", ylab = "Mathematical IQ")
abline(lm(wong2$piq ~ wong2$viq), col="blue")
linm1 <- lm(wong2$piq ~ wong2$viq)
summary(linm1)
cor(wong2$piq, wong2$viq)
```

STEP 2
```{r}
data <- read_excel("VIQT_data.xls")

#cutting the data of score_full distribution and multiplying by 1.2 to have the mean unchanged
data$score_full <- data$score_full*1.2
data0 <- subset(data, data$score_full>0)

#cutting the data to have the right age
data1 <- subset(data0, data0$age<100)

#expressing IQ from score by the formula score*100/age and cutting the data to have right distribution
data2 <- subset(data1, data1$score_full*100/data1$age<201)
```

STEP 3
In this code we grouped people by the age - before 20 - from 20 to 25 - from 25 to 30 - and up to 60 like this - so we have 9 groups
Then in each group we found the mean of IQ of people, who were in coma and of people who weren`t and found the differences between these means - You can see these differences plotted
```{r}
age <- c(25, 30, 35, 40, 45, 50, 55, 60)
patient <- vector()
people <- vector()
difference <- vector()
current_wong <- subset(wong2, wong2$age < 20)
patient_mean <- mean(current_wong$viq)
patient <- c(patient, patient_mean)
current_data <- subset(data2, data2$age < 20)
people_mean <- mean(current_data$score_full*100/current_data$age)
people <- c(people, people_mean)
difference <- c(difference, patient_mean - people_mean)
for(time in 1:length(age)){
  first <- age[time]
  second <- first - 5
  current_wong1 <- subset(wong2, wong2$age < first)
  current_wong <- subset(current_wong1, current_wong1$age >= second)
  patient_mean <- mean(current_wong$viq)
  patient <- c(patient, patient_mean)
  current_data <- subset(data2, data2$age < first)
  current_data <- subset(current_data, current_data$age >= second)
  people_mean <- mean(current_data$score_full*100/current_data$age)
  people <- c(people, people_mean)
  difference <- c(difference, patient_mean - people_mean)
}
age <- c(20, 25, 30, 35, 40, 45, 50, 55, 60)
plot(y = difference,x = age, type = "l", col = "blue")
```
STEP 4
```{r}
#cutting the data where people were in coma and their IQ was measures no longer from coma than about 3 years
second_data <- subset(first_data, first_data$duration>0)
second_data <- subset(second_data, second_data$days<1000)

#cutting the data and left there people, whose IQ was measured more than 1 time
rep_pat = data.frame(with(second_data, table(id)))
relevant_pat= subset(rep_pat , rep_pat$Freq>1)

data = subset(second_data, second_data$id %in% relevant_pat$id)
```

STEP 5
In this code we use the data with only patients, whose IQ was measured more than 1 time
We groupped people by the age - before 20 - from 20 to 25 - from 25 to 30 - and up to 60 like this - so we have 9 groups
1) For every patient we built the linear model - how her/his was recovering from coma - her/his IQ was measured several times during 3 years after getting out of a coma - how time influence on increasing of their IQ
2) We found the mean coefficiens  of this influence in each of 9 groups and plotted it

So on the plot You can see X - age, Y - how much time influences on increasing of IQ
```{r}
means <- vector()
g20 <- subset(data, data$age<20)
coef <- vector()
data1 <- g20[!duplicated(g20$id),]
for(n in 1:length(data1$id)){
  my_data <- subset(g20, g20$id == data1$id[n])
  model <- lm(my_data$viq ~ my_data$days)
  coefficient <- as.numeric(model$coefficients[2])
  coef = c(coef, coefficient)
}
means <- c(means, (mean(coef)))
age <- c(25, 30, 35, 40, 45, 50, 54, 60)
for(time in 1:length(age)){
  first <- age[time]
  second <- first - 5
  g201 <- subset(data, data$age<first)
  g20 <- subset(g201, g201$age>=second)
  coef <- vector()
  data1 <- g20[!duplicated(g20$id),]
  for(n in 1:length(data1$id)){
    my_data <- subset(g20, g20$id == data1$id[n])
    model <- lm(my_data$viq ~ my_data$days)
    coefficient <- as.numeric(model$coefficients[2])
    coef = c(coef, coefficient)
  }
  means <- c(means, (mean(coef)))
}
age <- c(20, 25, 30, 35, 40, 45, 50, 54, 60)
plot(y = means, x = age, type = "l", col= "blue")
```

STEP 6
On this plot You can see how how the rate of exit from a coma and the size of the brain injury depends
```{r}
plot(means, difference)
abline(lm(difference ~ means), col="blue")
linearmod <- lm(difference ~ means)
summary(linearmod)
```


SECTION 2

STEP 1
```{r}
library(dplyr)
library(LambertW)

data <- read_excel("VIQT_data.xls")

country_count = data.frame(with(data, table(country)))
relevant_country = subset(country_count, country_count$Freq>30)
#relevant country - country where were made more than 30 IQ tests

data = subset(data, data$country %in% relevant_country$country) #cut countries with less then 30 survey
data = subset (data, data$age > 5  & data$age< 80) #cut not relevant ages
```


STEP 2
```{r}
IQ = as.integer(data$score_full/data$age*100)
plot(data$age, IQ, xlab="age")
lines(lowess(data$age, IQ), col="blue") 
# Counting IQ by formula and draw its plot

plot( data$age, data$score_full, xlab="age", ylab="full score")
lines(lowess( data$age, data$score_full), col="blue") 
# Draw plot of dependence full score of age
```

STEP 3
```{r}
hist (IQ,  breaks=20)
# Our IQ- distribunion looks like normal distribution, now we have ti check it

qqnorm(IQ)
qqline(IQ, col = "black", lwd = 3)
# Draw QQline of our distribution and comrare it to normal distribution

shapiro.test(IQ[1:5000])
shapiro.test(IQ[5000:9915])
# Devide our data on two groups and make a sharpio test for each of them

ks.test.t(IQ)

#Sharpio-Wilk test and Kolmogorov-Smirnov give as p-value less then 2,2e-16 
# this mean that our data is perfectry Normally -distributed

print(summary(IQ))
print(sd(IQ))

#distribution of our IQ is  a normal distribution with mean 102 and sd 46, it is close to the real IQ distribution

data$IQ = IQ
# add column IQ to our dataframe
```

STEP 4
```{r}
country_IQ =aggregate(data, by=list(data$country),FUN=mean, na.rm=TRUE)[c(1,133)]

# We create a data_frame country_IQ with country and responsed mean IQ for it.

gdp= read.csv('cor_gdp.csv')
# cor_gdp.csv is a file with countries abbreviation and GDP per 2018 year

IQ_GDP = merge(country_IQ, gdp,by.x = 'Group.1' ,by.y = 'Country', all=F)
# We get intersect of country_IQ and gdp by Counrty

IQ_GDP = IQ_GDP %>% 
  rename(
    Country= Group.1
    )
# rename column 

print(IQ_GDP)
```

STEP 5
```{r}
plot(IQ_GDP$IQ, IQ_GDP$GDP, xlab="IQ", ylab = "GDP")
abline(lm(IQ_GDP$GDP~ IQ_GDP$IQ), col="blue")
# Now we don't see significant correlation

IQ_GDP$logged_GDP = log(IQ_GDP$GDP)

plot(IQ_GDP$IQ, IQ_GDP$logged_GDP, xlab="IQ", ylab = "logged_GDP")
abline(lm(IQ_GDP$logged_GDP~ IQ_GDP$IQ), col="blue")
# This way correlation is more clear because GDP is really big when average IQ doesn't have a bit range

cor(IQ_GDP$IQ, IQ_GDP$GDP)
cor(IQ_GDP$IQ, IQ_GDP$logged_GDP)
# Correlations equal to  0.3087967 and 0.4402467

print(summary(lm(IQ_GDP$GDP ~ IQ_GDP$IQ)))
print(summary(lm(IQ_GDP$logged_GDP ~ IQ_GDP$IQ)))

#Using lm we can see that GDP is statistically significant for IQ
```

STEP 6
```{r}
data = subset(data, data$gender ==2 | data$gender ==1) #cut third gender for relevant lm 

X <- model.matrix(~0+ engnat + education + urban + gender, data = data)
cor(X)
# We test our factors for dependence, aswe can see, variables are not dependent
# There is a slight correlation between urban and not native English, it means that most of the foreigners were taken from cities


print(summary(lm( IQ ~ engnat + education + urban + gender, data=data)))

# We can see, that engnat, education and gender impacts IQ 
# gender : boy's IQ is higher 
# people, whos native language was English passed the test much better
# and we can determinate that higher education level leads to lover IQ
# urbanization doesnt impact IQ
```

STEP 7
```{r}
X <- model.matrix(~0+S1+S2+S3+S4+S5+S6+S7+S8+S9+S10+S11+S12+S13+S14+S15+S16+S17+S18+S19+S20+S21+S22+S23+S24+S25+S26+S27+S28+S29+S30, data = data)
X=cor(X)
print(X)
# Let's test our questions for independense 
# Here is questions with correlation bigger, than 0,3 (20~26 = 0,41), (12~6 = 0,44), (2~4 = 0,5), (12~16 = 0,35) (14~13 = 0,33), (18~20 = 0,42)

model1 = lm(IQ~S1+S2+S3+S4+S5+S6+S7+S8+S9+S10+S11+S12+S13+S14+S15+S16+S17+S18+S19+S20+S21+S22+S23+S24+S25+S26+S27+S28+S29+S30, data= data)
step <- stepAIC(model1, direction="both", k=31)
step$anova # display results
# ve use variable selection to delete urrelevant variables from our model
# Final Model (only relevant):
#IQ ~ S3 + S6 + S8 + S9 + S10 + S11 + S12 + S13 + S15 + S16 + 
#    S17 + S18 + S19 + S20 + S21 + S22 + S23 + S25 + S26 + S27 + 
#    S28 + S29 + S30

model2 = lm (IQ ~  S3 + S9 + S10 + S12 + S13 + S18 + S19 + S20 + S21 + S22 +  S25 + S27 + S28 + S29 + S30, data = data)
print(summary(model2))

# As we can see, reliable questions are S3, S9 , S10 , S12 , S13 , S18 , S19 , S20 , S21 , S22 , S25 ,  S27 , S28 , S29 , S30 
# We exclude hightly correlated (26, 18, 6, 16, 14) and get S3, S9 , S10 , S12 , S19 , S20 , S21 , S22 , S25 ,S27 , S28 , S29 , S30 

# Here are some graffics of them
plot(data$S25, data$IQ, main ="I wish people were more spiritual", xlab="S9", ylab="IQ", ylim = c(0, 200))
lines(lowess(data$S25, data$IQ), col="blue") 

plot(data$S20, data$IQ, main ="I do more than someone expected of me.", xlab="S20", ylab="IQ", ylim = c(0, 200))
lines(lowess(data$S20, data$IQ), col="blue") 

plot(data$S30, data$IQ, main ="I am not quite sure what I want.", xlab="S30", ylab="IQ", ylim = c(0, 200))
lines(lowess(data$S30, data$IQ), col="blue") 

plot(data$S19, data$IQ, main ="I do not have a very expressive face.", xlab="S19", ylab="IQ", ylim = c(0, 200))
lines(lowess(data$S19, data$IQ), col="blue")
```